A numerical study of variational data assimilation for the wave equation in heterogeneous media

### 2025-06-24 tex/main.tex
Some polishing

\abstract{In recent years several numerical methods for solving the unique continuation problem for the wave equation in a homogeneous medium with given data on the lateral boundary of the space-time cylinder have been proposed. This problem enjoys Lipschitz stability if the geometric control condition is fulfilled, which allows to devise optimally convergent numerical methods. In this article we investigate whether these results carry over to the case in which the medium exhibits a jump discontinuity. Our numerical experiments suggest a positive answer. However, we also observe that the presence of discontinuities in the medium renders the computations far more demanding than in the homogeneous case.
}

@@

\section{Introduction}
\todo[inline]{@JP: please update affiliation}
\noindent We consider the discretization of a data assimilation problem for the wave equation in heterogeneous media. In the following, let $\Omega \subset \mathbb{R}^d$ be a bounded domain with smooth boundary $\partial \Omega$. We assume that $\Omega$ is split by a smooth $(d-1)$-dimensional 
interface $\Gamma$ into two disjoint components, i.e.\ $\Omega \setminus \Gamma = \Omega_1 \cup 
\Omega_2$ and $\Omega_1 \cap \Omega_2 = \emptyset$.
 We consider a scalar wave speed
\[
 c(x) := \begin{cases}
       c_1(x),  & x \in \Omega_1, \\
        c_2(x), & x \in \Omega_2,
    \end{cases}
    \quad 
    \text{ with } c_i \in C^\infty(\Omega_i), \; i = 1,2,
\]
 %$c(x) = \mathbbm{1}_{\Omega_1} c_1(x) + \mathbbm{1}_{\Omega_2} c_2(x)$, $c_i \in C^\infty(\Omega_i)$, $i = 1,2$,
which may be discontinuous across the interface $\Gamma$, but constant in time. For a final time $T > 0$, we consider the space-time domain $Q = (0,T) \times \Omega$ with lateral boundary $\Sigma \coloneqq (0,T) \times \partial \Omega$. For any nonempty subset $\omega \subset \Omega$, we set $\omega_T \coloneqq (0,T) \times \omega$. In the following, let $\wop$ be the wave operator with discontinuous wave-speed defined as 
\begin{equation*}
    \wop u := \partial_t^2 - \div(c^2 \nabla u).
\end{equation*}
In particular, we write $\square_1$ if $c_i = 1$, $i = 1,2$, i.e. if we are considering the case of homogeneous media. 
Then, we consider the problem: find $u : Q \rightarrow \mathbb{R}$ such that 
\begin{equation}\label{eq:waveEquation}
        (\wop u, u) = (0,0) \quad \text{ in } Q \times \Sigma,
\end{equation}
along with the natural interface conditions 
\begin{equation}\label{eq:IF-conditions}
	\jump{u}_{\Gamma} = 0 \text{ on } \Sigma, \qquad 
	\quad \jump{c^2 \nabla u}_{\Gamma} \cdot \mathbf{n}_{\Gamma} = 0 \text{ on } \Sigma, 
\end{equation}
where $\mathbf{n}_{\Gamma}$ is the normal vector on $\Gamma$ pointing into $\Omega_+$ 
and the jump across $\Gamma$ is defined\textsuperscript{1}\footnotemark \footnotetext{\textsuperscript{1}And anologously for vector-valued functions.} as $\jump{u}_{\Gamma} = u_{+}-u_{-}$ with $u_{\pm}$ being the traces 
of $u|_{\Omega_{\pm}}$ on $\Gamma$. \par 
\noindent To ensure that the problem is well posed, one usually prescribes initial data
\begin{equation}\label{eq:initialData}
    (u,\dt u) \vert_{t = 0} = (u_0,u_1) \text{ in } \Omega, \tag{IVP} 
\end{equation}
where $u_0 \in H^1(Q)$ and $u_1 \in L^2(Q)$ are given. For a detailed analysis of the well-posedness of the problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:initialData} in heterogeneous media we refer to \cite{StolkPhD}. 
Here we consider a data assimilation problem instead, i.e. we assume that the initial data is unknown and that there are given measurements $u_{\omega} \in L^2(\omega_T)$ which are prescribed in a data domain $\omega_T$, $\omega \subset \Omega$, $\omega \not = \emptyset$:
\begin{equation}\label{eq:dataMatch}
    u = u_{\omega} \text{ in } \omega_T. \tag{DA}
\end{equation}

%\noindent In a homogeneous medium, where $c_i = 1$, $i = 1,2$,
\noindent In a medium without an interface the problem \eqref{eq:waveEquation}+\eqref{eq:dataMatch} is well-studied, see e.g.~References \cite{CM15,BFO20,BFMO21,MM21,DMS23,BDE24}. In particular, it is known that if $c_i = 1$, $i = 1,2$, the boundary $\partial \Omega$ is strictly convex and the set $\omega_T \subset Q$ fulfills the geometric control condition (GCC) \cite{BLR92}, then the problem is Lipschitz stable. Roughly speaking, a set $\omega \subset \Omega$ and a final time $T$ fulfill the GCC if all geometric optic rays traveling with unit speed\textsuperscript{2}\footnotemark \footnotetext{\textsuperscript{2}with adapting the metric to the wave speed $c$.} in $\Omega$, taking into account possible reflections on the boundary $\partial \Omega$, intersect the set $(0,T) \times \omega$. For a precise definition, we refer to \cite{BLR92}. Assuming that the GCC is satisfied, we have the following result.

\begin{thm}\label{thm:Lipschitz}
    Assume that $c_i = 1$, $i = 1,2$, and that $\partial \Omega$ is stricly convex. If $\omega_T \subset Q$ fulfills the GCC, then there exists a constant $C > 0$ such that for any $\phi \in H^1(Q)$ we have the following estimates
    \begin{align*}
        \Vert \phi \Vert_{L^\infty(0,T;L^2(\Omega))} + \Vert \dt \phi \Vert_{L^2(0,T;H^{-1}(\Omega))} &\le C \left(  \Vert \phi \Vert_{L^2(\omega_T)} + \Vert \phi \Vert_{L^2(\Sigma)} + \Vert \square_1 \phi \Vert_{H^{-1}(Q)} \right), \\
        \Vert \phi \vert_{t = 0} \Vert_{L^2(\Omega)} + \Vert \dt \phi \vert_{t = 0} \Vert_{H^{-1}(\Omega)} &\le C \left(\Vert \phi \Vert_{L^2(\omega_T)} + \Vert \phi \Vert_{L^2(\Sigma)} + \Vert \square_1 \phi \Vert_{H^{-1}(Q)} \right). 
    \end{align*}
\end{thm}

@@

\noindent For the heterogeneous case where $c_1 \not =  c_2$, the problem is less studied. In the absence of the GCC, we have to assume that the final time $T$ is large enough to ensure that problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} is well-posed. To be precise, Holmgren's unique continuation theorem guarantees that \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} is uniquely solvable if
\begin{equation}\label{eq:Tcondition}
    T > 2 \sup_{x \in \Omega} \operatorname{dist}_{c}(x,\omega_T),
\end{equation}
where $\operatorname{dist}_c(x,\omega_T) := \inf_{y \in \omega_T} d_c(x,y)$ with $d_c$ being the distance function defined as the length of the shortest continuous path in $\Omega$ connecting $x$ and $y$ with respect to the metric adapted to the discontinuous wave speed $c$. While the condition \eqref{eq:Tcondition} guarantees unique solvability, the stability of problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} can be poor. It has been shown in \cite[Thm. 5.19]{Filippas22} that under the assumption \eqref{eq:Tcondition} the following stability result holds:  
\begin{thm}[Thm. 5.19 of \cite{Filippas22}]\label{thm:filippas}
    Let $\omega_T \subset \Omega$ be nonempty and let $T$ be s.t. \eqref{eq:Tcondition} holds. Then, there exists $C,\kappa$ such that for any initial data $(u_0,u_1) \in H^1_0(\Omega) \times L^2(\Omega)$ and $u$ solving \eqref{eq:IF-conditions}+\eqref{eq:initialData} and  
\[
\wop u \in L^2(Q), \quad u = 0 \text{ on } \Sigma,
\]
one has for any $\mu > 0$ that 
    \begin{equation}\label{eq:FilippasEstimate}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C e^{\kappa \mu} \left( \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)} \right) + \frac{C}{\mu} \Vert (u_0,u_1) \Vert_{H^1 \times L^2}. 
    \end{equation}
    In particular, if $(u_0,u_1) \not = (0,0)$ we have that 
    \begin{equation}\label{eq:FilippasEstimateLog}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{\log \left( 1 + \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{ \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)}} \right)}. 
    \end{equation} 
\end{thm}

\noindent In this article, we introduce a discontinuous in time finite element method to solve the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} in the heterogeneous case. In particular, we want to investigate numerically if the following assumption, motivated by Thm. \ref{thm:Lipschitz} for the homogeneous case, is justified.

@@

That is, we would like to understand whether the Lipschitz stability property is preserved in presence of strong heterogeneity in the medium provided that the GCC holds. This question is of high practical importance since discontinuous media appear in many applications, e.g.\ in seismology. If the corresponding data assimilation problems were indeed subject to the poor logarithmic stability property ensured by \eqref{eq:FilippasEstimateLog}, this would pose an enormous challenge for the design of accurate and realiable numerical methods for their resolution.

\section{Discretization (fully discrete \& dG in time)} 
\noindent In this section, we introduce a fully discrete discontinuous in time finite element method to discretize the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} for the wave equation in heterogeneous media. The discretization is based on the scheme introduced in \cite{BP24} for the homogeneous case. Before introducing the method in Section \ref{sec:method}, we describe the geometric partition of the space-time domain $Q$ into time slabs in Section \ref{sec:spaceTimeDiscretization}.

\subsection{Discretization of the space-time domain}\label{sec:spaceTimeDiscretization}
Let $\hat{\mathcal{T}}_h$ be a quasi-uniform triangulation with mesh size $h$ such that $\hat{\Omega}:= \cup_{K \in \hat{\mathcal{T}}_h} K $ contains $\Omega$, i.e.\ $\Omega \subset \hat{\Omega}$. Setting $\mathcal{T}_h := \{ K \cap \Omega \mid K \in  \hat{\mathcal{T}}_h \}$ then ensures that $\Omega = \cup_{K \in \mathcal{T}_h} K$ holds true.  
We recall from  \cite[Sec. 4.2]{BFMO21control} that the triangulation $\{ \hat{\mathcal{T}}_h \mid h > 0 \}$ can be constructed such that for $h$ small enough the following continuous trace inequality holds:  For all $v \in [H^1(K)]^d$ and $K \in \mathcal{T}_h$, we have
\begin{equation}\label{eq:traceInequality}
    \Vert v \Vert_{[L^2(\partial K)]^d} \le C \left(h^{-1/2} \Vert v \Vert_{[L^2(K)]^d} + h^{1/2} \Vert \nabla v \Vert_{[L^2(K)]^d} \right).  
\end{equation}
We assume in what follows additionally that the triangulation $\mathcal{T}_h$ fits the interface $\Gamma$ and the data domain $\omega$. An extension to allow for unfitted interfaces could be possible utilizing techniques we introduced for a stationary problem in \cite{BP25}, but is certainly beyond the scope of this article.

\noindent For $k \ge 1$, we define the $H^1$-conforming finite element space 
\begin{equation}
    V_h^k := \{ v \in H^1(\Omega) : v \vert_{K} \in \mathcal{P}^k(K) \ \forall K \in \mathcal{T}_h \},
\end{equation}
where $\mathcal{P}^k(K)$ denotes the space of polynomials of degree at most $k \in \mathbb{N}$ on $K \in \mathcal{T}_h$. Furthermore, we partition the time axis into $N$ subintervals $I_n = (t_n,t_{n+1})$, $n = 0, \dots, N-1$, where $0 = t_0 \le t_1 \le \dots \le t_N = T$. We assume that the intervals are of equal length and denote $\Delta t = \vert t_{n+1} -t_n \vert$. Then, we partition $Q$ and $\Sigma$ into time-slabs 
\begin{equation}
    \begin{aligned}
        Q^n := &I_n \times \Omega, \quad \Sigma^n := I_n \times \Sigma, \quad n = 0, \dots, N-1, \\
        Q = &\bigcup_{n = 0}^{N-1} Q^n, \quad \phantom{:} \Sigma = \bigcup_{n = 0}^{N-1} \Sigma^n.
    \end{aligned}
\end{equation}
In the following, we denote the space-time integrals on the time slabs as 
\begin{equation*}
    (u,v)_{Q^n} := \int_{I_n} \int_{\Omega} uv \ \dX \dT, \quad (u,v)_{\Sigma^n} := \int_{I_n} \int_{\Sigma} uv \ \dS \dT,
\end{equation*}
and define $\Vert v \Vert^2_{Q^n} := (v,v)_{Q^n}$ and $\Vert v \Vert^2_{\Sigma^n} := (v,v)_{\Sigma^n}$. 
Furthermore, we set 
\begin{equation*}
    \omega^n := I_n \times \omega, \quad (u,v)_{\omega^n} := \int_{I_n} \int_{\omega} uv \ \dX \dT, \quad \Vert v \Vert^2_{\omega_T} := \sum_{n = 0}^{N-1} (v,v)_{\omega^n}.
\end{equation*}
Finally, we define the time jump operator 
\begin{equation*}
    v^n_{\pm} (x) := \lim_{s \rightarrow 0^+} v(x,t_n \pm s), \quad \jump{v^n} := v^n_+ - v^n_-. 
\end{equation*}

\subsection{A discontinuous in time FEM for the data assimilation problem}\label{sec:method}
\noindent We define the discontinuous in time finite element spaces 

@@

~\jp{I think that in the CIP we have to switch from the full gradient to the normal derivative  $\jump{c^2 \nabla \ul_1} \cdot \mathbf{n}$ only. Indeed, it follows from the first condition in \eqref{eq:IF-conditions} that the tangential part of the gradient on $\Gamma$ must be continuous $\jump{\nabla_{\Gamma} u} = 0$. So stabbilizing the full gradient is not consistent, which becomes an issue in Thm.~\ref{thm:bestapprox} as $S_h(\mathbf{U},\Wh) = 0$ would not hold.

@@

\noindent With the definition of $\cL_h$, the first order optimality conditions take the form: Find $(\Uh,\Zh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}$ such that 
\begin{alignat*}{2}
    (\ul_1,\wl_1)_{\omega_T} \! + \! A[\Wh,\Zh] + S_h(\Uh,\Wh) + \Sud(\Uh,\Wh) &= (u_{\omega},\wl_1)_{\omega_T} \ &&\forall \Wh \in \ProdFullyDiscrSpace{k}{q} \\
    A[\Uh,\Yh] -  S_h^\ast(\Yh,\Zh) &= 0 \ &&\forall \Yh \in \ProdFullyDiscrSpace{k^\ast}{q^\ast} 
\end{alignat*}
In a more compact form, we can write these conditions as: Find $(\Uh,\Zh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}$ such that
\begin{equation}\label{eq:discreteProblem}
    B[(\Uh,\Zh),(\Wh,\Yh)] = (u_{\omega},\wl_1)_{\omega_T} \quad \forall (\Wh,\Yh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast},
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        B[(\Uh,\Zh),(\Wh,\Yh)] \coloneqq \ &(\ul_1,\wl_1)_{\omega_T} + A[\Wh,\Zh]+ S_h(\Uh,\Wh) \\
        &+  \Sud(\Uh,\Wh) + A[\Uh,\Yh] - S_h^\ast(\Yh,\Zh).
    \end{aligned}
\end{equation}

>>>>

\abstract{In recent years, several numerical methods for solving the unique continuation problem for the wave equation in a homogeneous medium with given data on the lateral boundary of the space-time cylinder have been proposed. This problem enjoys Lipschitz stability if the geometric control condition is fulfilled, which allows devising optimally convergent numerical methods. In this article, we investigate whether these results carry over to the case in which the medium exhibits a jump discontinuity. Our numerical experiments suggest a positive answer. However, we also observe that the presence of discontinuities in the medium renders the computations far more demanding than in the homogeneous case.
}

@@

\section{Introduction}
\todo[inline]{@JP: please update affiliation}
\todo[inline]{@JP: Let's discuss the title again, I feel like we should include 'stability' somewhere, because it's our main objective...}
We consider the discretization of a data assimilation problem for the wave equation in heterogeneous media. In the following, let $\Omega \subset \mathbb{R}^d$ be a bounded domain with smooth boundary $\partial \Omega$. We assume that $\Omega$ is split by a smooth $(d-1)$-dimensional interface $\Gamma$ into two disjoint components such that $\Omega \setminus \Gamma = \Omega_1 \cup 
\Omega_2$ and $\Omega_1 \cap \Omega_2 = \emptyset$.

We consider a scalar wave speed
\[
 c(x) := \begin{cases}
       c_1(x),  & x \in \Omega_1, \\
        c_2(x), & x \in \Omega_2,
    \end{cases}
    \quad 
    \text{ with } c_i \in C^\infty(\Omega_i), \; i = 1,2,
\]
 %$c(x) = \mathbbm{1}_{\Omega_1} c_1(x) + \mathbbm{1}_{\Omega_2} c_2(x)$, $c_i \in C^\infty(\Omega_i)$, $i = 1,2$,
which may be discontinuous across the interface $\Gamma$, but is constant in time. For a final time $T > 0$, we consider the space-time domain $Q = (0,T) \times \Omega$ with lateral boundary $\Sigma \coloneqq (0,T) \times \partial \Omega$, and for any nonempty subset $\omega \subset \Omega$, we set $\omega_T \coloneqq (0,T) \times \omega$.

Let $\wop$ be the wave operator with discontinuous wave-speed defined as 
\begin{equation*}
    \wop u := \partial_t^2 u - \div(c^2 \nabla u).
\end{equation*}
In particular, we write $\square_1$ if $c_i = 1$ for $i = 1,2$, corresponding to the case of a homogeneous medium with unit wave speed.

Then, we consider the problem: find $u : Q \rightarrow \mathbb{R}$ such that 
\begin{equation}\label{eq:waveEquation}
        (\wop u, u) = (0,0) \quad \text{ in } Q \times \Sigma,
\end{equation}
along with the natural interface conditions 
\begin{equation}\label{eq:IF-conditions}
	\jump{u}_{\Gamma} = 0 \text{ on } \Sigma, \qquad 
	\quad \jump{c^2 \nabla u}_{\Gamma} \cdot \mathbf{n}_{\Gamma} = 0 \text{ on } \Sigma, 
\end{equation}
where $\mathbf{n}_{\Gamma}$ is the normal vector on $\Gamma$ pointing into $\Omega_+$ 
and the jump across $\Gamma$ is defined\textsuperscript{1}\footnotemark \footnotetext{\textsuperscript{1}And analogously for vector-valued functions.} as $\jump{u}_{\Gamma} = u_{+}-u_{-}$ with $u_{\pm}$ being the traces 
of $u|_{\Omega_{\pm}}$ on $\Gamma$. \par 
To ensure that the problem is well posed, one usually prescribes initial data
\begin{equation}\label{eq:initialData}
    (u,\dt u) \vert_{t = 0} = (u_0,u_1) \text{ in } \Omega, \tag{IVP} 
\end{equation}
where $u_0 \in H^1(Q)$ and $u_1 \in L^2(Q)$ are given. For a detailed analysis of the well-posedness of the problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:initialData} in heterogeneous media, we refer to \cite{StolkPhD}.

In this work, we consider a \emph{data assimilation} problem instead, where the initial data is unknown, and we assume that there are given measurements $u_{\omega} \in L^2(\omega_T)$ prescribed in a data domain $\omega_T$, with $\omega \subset \Omega$, $\omega \not = \emptyset$:
\begin{equation}\label{eq:dataMatch}
    u = u_{\omega} \text{ in } \omega_T. \tag{DA}
\end{equation}
In a medium without an interface, the problem \eqref{eq:waveEquation}+\eqref{eq:dataMatch} is well-studied; see for example \cite{CM15,BFO20,BFMO21,MM21,DMS23,BDE24}. In particular, if the boundary $\partial \Omega$ is strictly convex, the problem is Lipschitz stable provided that the set $\omega_T \subset Q$ fulfills the geometric control condition (GCC) \cite{BLR92}. 
Roughly speaking, a set $\omega \subset \Omega$ and a final time $T$ fulfill the GCC if all geometric optic rays traveling with unit speed\textsuperscript{2}\footnotemark \footnotetext{\textsuperscript{2}We consider the metric to be adapted to the wave speed $c$.} in $\Omega$, accounting for possible reflections on the boundary $\partial \Omega$, intersect the set $(0,T) \times \omega$. For a more precise definition, we refer to \cite{BLR92}. Assuming that the GCC is satisfied, we have the following result.

\begin{thm}\label{thm:Lipschitz}
    Assume that $\partial \Omega$ is strictly convex and $c_i = 1$ for $i = 1,2$. If $\omega_T \subset Q$ fulfills the GCC, then there exists a constant $C > 0$ such that for any $\phi \in H^1(Q)$, we have the following estimates:
    \begin{align*}
        \Vert \phi \Vert_{L^\infty(0,T;L^2(\Omega))} + \Vert \dt \phi \Vert_{L^2(0,T;H^{-1}(\Omega))} &\le C \left(  \Vert \phi \Vert_{L^2(\omega_T)} + \Vert \phi \Vert_{L^2(\Sigma)} + \Vert \square_1 \phi \Vert_{H^{-1}(Q)} \right), \\
        \Vert \phi \vert_{t = 0} \Vert_{L^2(\Omega)} + \Vert \dt \phi \vert_{t = 0} \Vert_{H^{-1}(\Omega)} &\le C \left(\Vert \phi \Vert_{L^2(\omega_T)} + \Vert \phi \Vert_{L^2(\Sigma)} + \Vert \square_1 \phi \Vert_{H^{-1}(Q)} \right). 
    \end{align*}
\end{thm}

@@

In the heterogeneous case, where $c_1 \not =  c_2$, the problem is less studied. In the absence of the GCC, we have to assume that the final time $T$ is large enough to ensure that problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} is well-posed. To be precise, Holmgren's unique continuation theorem guarantees that \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} is uniquely solvable if
\begin{equation}\label{eq:Tcondition}
    T > 2 \sup_{x \in \Omega} \operatorname{dist}_{c}(x,\omega_T),
\end{equation}
where $\operatorname{dist}_c(x,\omega_T) := \inf_{y \in \omega_T} d_c(x,y)$. Here, $d_c$ is the distance function defined as the length of the shortest continuous path in $\Omega$ connecting $x$ and $y$ with respect to the metric adapted to the discontinuous wave speed $c$. While the condition \eqref{eq:Tcondition} guarantees unique solvability, the stability of problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} can be poor. However, as shown \cite[Thm. 5.19]{Filippas22}, assumption \eqref{eq:Tcondition} grants the following stability result holds.

\begin{thm}[Thm. 5.19 of \cite{Filippas22}]\label{thm:filippas}
    Let $\omega_T \subset \Omega$ be nonempty and let $T > 0$ be s.t. \eqref{eq:Tcondition} holds. Then, there exists $C,\kappa$ such that for any initial data $(u_0,u_1) \in H^1_0(\Omega) \times L^2(\Omega)$ and $u$ solving \eqref{eq:IF-conditions}+\eqref{eq:initialData} and  
\[
\wop u \in L^2(Q), \quad u = 0 \text{ on } \Sigma,
\]
one has for any $\mu > 0$ that 
    \begin{equation}\label{eq:FilippasEstimate}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C e^{\kappa \mu} \left( \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)} \right) + \frac{C}{\mu} \Vert (u_0,u_1) \Vert_{H^1 \times L^2}. 
    \end{equation}
    In particular, if $(u_0,u_1) \not = (0,0)$ we have that 
    \begin{equation}\label{eq:FilippasEstimateLog}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{\log \left( 1 + \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{ \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)}} \right)}. 
    \end{equation} 
\end{thm}

\todo[inline]{@JP: I made some minor changes to the introduction... Here, I felt that we should manage the reader's expectation a bit more and stress that we do not focus on a complete analysis. }

In this article, we consider a discontinuous-in-time finite element method to solve the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} in the heterogeneous case. This method was originally introduced for the homogeneous case in \cite{BP24}. The focus of this article is not on the numerical analysis of the proposed discretization, but rather on investigating the numerically observed stability under the assumption of the GCC. Specifically, we want to investigate numerically whether the following assumption -- motivated by \cref{thm:Lipschitz} for the homogeneous case -- is justified.

@@

In summary, our goal is to understand whether the Lipschitz stability property is preserved in the presence of strong heterogeneity in the medium, provided that the GCC holds. This question is of high practical importance, as discontinuous media appear in many applications, for example, in seismology. If the corresponding data assimilation problems were indeed subject to the poor logarithmic stability property ensured by \eqref{eq:FilippasEstimateLog}, this would pose an enormous challenge for the design of accurate and reliable numerical methods.

\vspace*{1cm}
\noindent \textbf{Struture of the article.} In \cref{sec:discretization}, we introduce the method used to discretize \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch}.

\section{Fully discrete discontinuous-in-time discretization}\label{sec:discretization} 
In this section, we present a fully discrete discontinuous-in-time finite element method for discretizing the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch}, extending the method from \cite{BP24} to the heterogeneous case. \Cref{sec:spaceTimeDiscretization} describes the geometric partition of the space-time domain $Q$ into time slabs, followed by the full discretization in \cref{sec:method}.

\subsection{Discretization of the space-time domain}\label{sec:spaceTimeDiscretization}
Let $\hat{\mathcal{T}}_h$ be a quasi-uniform triangulation with mesh size $h$ such that $\hat{\Omega}:= \cup_{K \in \hat{\mathcal{T}}_h} K $ contains $\Omega$, i.e.\ $\Omega \subset \hat{\Omega}$. Setting $\mathcal{T}_h := \{ K \cap \Omega \mid K \in  \hat{\mathcal{T}}_h \}$ then ensures that $\Omega = \cup_{K \in \mathcal{T}_h} K$ holds true.  
We recall from  \cite[Sec. 4.2]{BFMO21control} that the triangulation $\{ \hat{\mathcal{T}}_h \mid h > 0 \}$ can be constructed such that for $h$ small enough the following continuous trace inequality holds:  For all $v \in [H^1(K)]^d$ and $K \in \mathcal{T}_h$, we have
\begin{equation}\label{eq:traceInequality}
    \Vert v \Vert_{[L^2(\partial K)]^d} \le C \left(h^{-1/2} \Vert v \Vert_{[L^2(K)]^d} + h^{1/2} \Vert \nabla v \Vert_{[L^2(K)]^d} \right).  
\end{equation}
We assume additionally that the triangulation $\mathcal{T}_h$ fits the interface $\Gamma$ and the data domain $\omega$. An extension to allow for unfitted interfaces could be possible by utilizing techniques introduced in \cite{BP25} for a stationary problem, but this is certainly beyond the scope of this article.

For polynomial degree $k \ge 1$, we define the $H^1$-conforming finite element space 
\begin{equation}
    V_h^k := \{ v \in H^1(\Omega) : v \vert_{K} \in \mathcal{P}^k(K) \ \forall K \in \mathcal{T}_h \},
\end{equation}
where $\mathcal{P}^k(K)$ denotes the space of polynomials of degree at most $k \in \mathbb{N}$ on $K \in \mathcal{T}_h$. Furthermore, we partition the time axis into $N$ subintervals $I_n = (t_n,t_{n+1})$, $n = 0, \dots, N-1$, where $0 = t_0 \le t_1 \le \dots \le t_N = T$. We assume that the intervals are of equal length and denote $\Delta t = \vert t_{n+1} -t_n \vert$. Then, we partition $Q$ and $\Sigma$ into time-slabs 
\begin{equation}
    \begin{aligned}
        Q^n := &I_n \times \Omega, \quad \Sigma^n := I_n \times \Sigma, \quad n = 0, \dots, N-1, \\
        Q = &\bigcup_{n = 0}^{N-1} Q^n, \quad \phantom{:} \Sigma = \bigcup_{n = 0}^{N-1} \Sigma^n.
    \end{aligned}
\end{equation}
In the following, we denote the space-time integrals on the time slabs as 
\begin{equation*}
    (u,v)_{Q^n} := \int_{I_n} \int_{\Omega} uv \ \dX \dT, \quad (u,v)_{\Sigma^n} := \int_{I_n} \int_{\Sigma} uv \ \dS \dT,
\end{equation*}
and define $\Vert v \Vert^2_{Q^n} := (v,v)_{Q^n}$ and $\Vert v \Vert^2_{\Sigma^n} := (v,v)_{\Sigma^n}$. 
Furthermore, we set 
\begin{equation*}
    \omega^n := I_n \times \omega, \quad (u,v)_{\omega^n} := \int_{I_n} \int_{\omega} uv \ \dX \dT, \quad \Vert v \Vert^2_{\omega_T} := \sum_{n = 0}^{N-1} (v,v)_{\omega^n}.
\end{equation*}
Finally, we define the time jump operator 
\begin{equation*}
    v^n_{\pm} (x) := \lim_{s \rightarrow 0^+} v(x,t_n \pm s), \quad \jump{v^n} := v^n_+ - v^n_-. 
\end{equation*}

\subsection{Full discretization}\label{sec:method}
 We define the discontinuous in time finite element spaces 

@@

~\jp{I think that in the CIP we have to switch from the full gradient to the normal derivative  $\jump{c^2 \nabla \ul_1} \cdot \mathbf{n}$ only. Indeed, it follows from the first condition in \eqref{eq:IF-conditions} that the tangential part of the gradient on $\Gamma$ must be continuous $\jump{\nabla_{\Gamma} u} = 0$. So stabbilizing the full gradient is not consistent, which becomes an issue in \cref{thm:bestapprox} as $S_h(\mathbf{U},\Wh) = 0$ would not hold.

@@

 With the definition of $\cL_h$, the first order optimality conditions take the form: Find $(\Uh,\Zh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}$ such that 
\begin{alignat*}{2}
    (\ul_1,\wl_1)_{\omega_T} \! + \! A[\Wh,\Zh] + S_h(\Uh,\Wh) + \Sud(\Uh,\Wh) &= (u_{\omega},\wl_1)_{\omega_T} \ &&\forall \Wh \in \ProdFullyDiscrSpace{k}{q} \\
    A[\Uh,\Yh] -  S_h^\ast(\Yh,\Zh) &= 0 \ &&\forall \Yh \in \ProdFullyDiscrSpace{k^\ast}{q^\ast} 
\end{alignat*}
In a more compact form, we can write these conditions as: Find $(\Uh,\Zh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}$ such that
\begin{equation}\label{eq:discreteProblem}
    B[(\Uh,\Zh),(\Wh,\Yh)] = (u_{\omega},\wl_1)_{\omega_T} \quad \forall (\Wh,\Yh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast},
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        B[(\Uh,\Zh),(\Wh,\Yh)] \coloneqq \ &(\ul_1,\wl_1)_{\omega_T} + A[\Wh,\Zh]+ S_h(\Uh,\Wh) \\
        &+  \Sud(\Uh,\Wh) + A[\Uh,\Yh] - S_h^\ast(\Yh,\Zh).
    \end{aligned}
\end{equation}

\section{Error analysis}
This section outlines how the steps from \cite{BP24} can be adapted to the heterogeneous setting to analyze the approximation of \eqref{eq:waveEquation}-\eqref{eq:IF-conditions}+\eqref{eq:dataMatch} through \eqref{eq:discreteProblem}. In \cref{sec:analysis:infsup}, we establish inf-sup stability and continuity of the discrete problem with respect to suitable discrete norms.
In \cref{sec:analysis:approx}, we show that the error between the first component $\ul_1$ of the discrete solution and the continuous solution $u$ is bounded by the best approximation error in these norms. Finally, we argue that \cref{assum:LipschitzStability} allows recovering convergence rates in a physically relevant norm.
